Making Deep Q-learning Methods Robust to Time Discretization

### 2019-01-30 paper/icml_drau/advup.tex

%\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)}
Still, this reparameterization does not, on its own, guarantee that $A$
correctly approximates $A^\pi_\deltat$ if
$Q_\Theta$ approximates $Q^\pi_\deltat$% \footnote{
% 	Ensuring that $A$ does not differ from $A^\pi_\deltat$ by a state dependent function only affects
% 	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
% 	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is
% 	the same. \TODO{still unclear why we'd care about identifiability}
%}
.
Indeed, for any given pair $(V_\theta,\,A_\psi)$, the pair $(V_\theta(s) -
f(s),\,A_\psi(s,a)+f(s)/\deltat)$ (for an arbitrary $f$)
% \begin{align}
% 	\tilde{V}(s) &= V(s) - f(s),\qquad
% 	\tilde{A}(s, a) = A(s, a) + \frac{f(s)}{\deltat}
% \end{align}
yields the exact same function $Q_\Theta$. This new $A_\psi$ still defines %the same $Q$ and
the same ranking of actions, yet this phenomenon might cause
numerical problems or instability of $A_\psi$ when $\deltat\to 0$, and prevents direct
interpretation of the learned $A_\psi$.
% Consequently, $Q$ can correctly approximate
% $Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
To enforce identifiability of $A_\psi$, one must enforce the consistency equation
\begin{equation}
	V^\pi_\deltat(s) = Q^\pi_\deltat(s,
	\pi(s))
\end{equation}
on the approximate $A_\psi$ and $V_\theta$. This translates to
\begin{equation}
	A_\psi(s, \pi(s)) = 0.
\end{equation}
With this additional constraint, if $Q_\Theta = Q^\pi_\deltat$, then $A_\psi =
A^\pi_\deltat$ and $V_\theta = V^\pi_\deltat$: indeed
\begin{align}
	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
		    &= \frac{Q_\Theta(s,a) - Q_\Theta(s, \pi(s))}{\deltat}
		    %\\ &
		    = A_\psi(s, a).
\end{align}
In the spirit of~\cite{dueling_nets}, instead of directly parameterizing $A_\psi$,
we define a parametric function $\bar{A}_\psi$ (typically a neural network),
and use $\bar{A}_\psi$ to define $A_\psi$ as
\begin{equation}
\label{eq:Aparam}
	A_\psi(s, a) \deq \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s)).
\end{equation}
% Rq Harsh
so that $A_\psi$ directly verifies the consistency condition.

@@

We learn $V_{\theta}$ and $A_{\psi}$ via suitable variants of $Q$-learning for continuous and
discrete action spaces. Namely, 
the true $A^\pi_\deltat$
and $V^\pi_\deltat$ of a near-continuous MDP with greedy exploitation
policy $\pi(s) \deq
\text{argmax}_{a'}A^\pi_\deltat(s, a')$
are the unique solution to the Bellman and consistency equations
\begin{align}
	V^{\pi}_\deltat(s) + \deltat\, A^{\pi}_\deltat(s, a) &=
	r\,\deltat + \gamma^{\deltat}  \,\E_{s'} V^{\pi}_\deltat(s')\label{eq:bellman_A}\\
	A^{\pi}_\deltat(s, \pi(s)) &= 0\label{eq:max_A}.
\end{align}
as seen in \ref{subsec:reparam}. Then $V_{\theta}$
and $A_{\psi}$ are trained to approximately solve these equations.

Maximization over actions in $\pi$ is implemented exactly for discrete actions,
and approximated by a policy neural network $\pi_\phi(s)$, trained to maximize $A_\psi(s,
\pi_\phi(s))$ for continuous actions, similarly to
\cite{ddpg}.

Eq.~\eqref{eq:max_A}
is directly verified by $A_{\psi}$, owing to the reparametrization
$A_\psi(s, a) = \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s))$, described
in~\ref{subsec:reparam}.  To approximately verify~\eqref{eq:bellman_A}, the
corresponding squared residual is minimized by an approximate gradient descent.
The update equations when learning from a transition $(s, a, r, s')$, either from
an exploratory trajectory or from a replay buffer~\cite{dqn}, are
\begin{align}
	\delta Q_\deltat &\leftarrow A_\psi(s, a)\, \deltat - \left(r
	\,\deltat + \gamma^{\deltat}\, V_\theta(s') - V_\theta(s)\right)
	\label{eq:approx_deltaQ}\\
	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat
	\,\partial_{\theta} V_\theta(s) \,\frac{\delta Q_\deltat}{\deltat}
	\label{eq:approx_bellman_V}\\
	\psi_\deltat &\leftarrow \psi_\deltat + \eta^A_\deltat
	\,\partial_{\psi} A_\psi(s, a) \,\frac{\delta Q_\deltat}{\deltat}.
	\label{eq:approx_bellman_A}
\end{align} 
where the $\eta$'s are learning rates.
Appropriate scalings for the learning rates $\eta^V_\deltat$ and
$\eta^A_\deltat$ in term of $\deltat$ to obtain a well defined continuous
limit are derived next.
% As we will see in the following section,
% to obtain a well defined continuous time limit, the learning rates can be rewritten as
% $\eta^i_\deltat = \alpha^i \deltat$, where $\alpha^1$ and $\alpha^2$ are independant
% of $\deltat$.

@@

Informally, in the parameter updates
\eqref{eq:approx_deltaQ}--\eqref{eq:approx_bellman_A}, the quantity $\delta
Q_\deltat$ is of order $O(\deltat)$, because $s'=s+O(\deltat)$ in a
near-continuous system. Therefore, $\delta
Q_\deltat/\deltat$ is $O(1)$, so that the gradients used to
update $\theta$ and $\psi$ are $O(1)$. Therefore, if the
learning rates are of order $\deltat$, one would expect 
the parameters $\theta$ and $\psi$ to move by $O(\deltat)$ in
each time interval $\deltat$, thus hopefully converging to smooth
continuous-time trajectories. The next theorem formally confirms that
learning rates of order $\deltat$ are the only possibility.

>>>>

%\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)}
Still, this reparameterization does not, on its own, guarantee that $A$
correctly approximates $A^\pi_\deltat$ if
$Q_\Theta$ approximates $Q^\pi_\deltat$% \footnote{
% 	Ensuring that $A$ does not differ from $A^\pi_\deltat$ by a state dependent function only affects
% 	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
% 	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is
% 	the same. \TODO{still unclear why we'd care about identifiability}
%}
.
Indeed, for any given pair $(V_\theta,\,A_\psi)$, the pair $(V_\theta(s) -
f(s),\,A_\psi(s,a)+f(s)/\deltat)$ (for an arbitrary $f$)
% \begin{align}
% 	\tilde{V}(s) &= V(s) - f(s),\qquad
% 	\tilde{A}(s, a) = A(s, a) + \frac{f(s)}{\deltat}
% \end{align}
yields the exact same function $Q_\Theta$. This new $A_\psi$ still defines %the same $Q$ and
the same ranking of actions, yet this phenomenon might cause
numerical problems or instability of $A_\psi$ when $\deltat\to 0$, and prevents direct
interpretation of the learned $A_\psi$.
% Consequently, $Q$ can correctly approximate
% $Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
To enforce identifiability of $A_\psi$, one must enforce the consistency equation
\begin{equation}
	V^\pi_\deltat(s) = Q^\pi_\deltat(s,
	\pi(s))
\end{equation}
on the approximate $A_\psi$ and $V_\theta$. This translates to
\begin{equation}
	A_\psi(s, \pi(s)) = 0.
\end{equation}
With this additional constraint, if $Q_\Theta = Q^\pi_\deltat$, then $A_\psi =
A^\pi_\deltat$ and $V_\theta = V^\pi_\deltat$: indeed
\begin{align}
	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
		    &= \frac{Q_\Theta(s,a) - Q_\Theta(s, \pi(s))}{\deltat}
		    %\\ &
		    = A_\psi(s, a).
\end{align}
In the spirit of~\cite{dueling_nets}, instead of directly parameterizing $A_\psi$,
we define a parametric function $\bar{A}_\psi$ (typically a neural network),
and use $\bar{A}_\psi$ to define $A_\psi$ as
\begin{equation}
\label{eq:Aparam}
	A_\psi(s, a) \deq \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s))
\end{equation}
% Rq Harsh
so that $A_\psi$ directly verifies the consistency condition.

@@

We learn $V_{\theta}$ and $A_{\psi}$ via suitable variants of $Q$-learning for continuous and
discrete action spaces. Namely, 
the true $A^\pi_\deltat$
and $V^\pi_\deltat$ of a near-continuous MDP with greedy exploitation
policy $\pi(s) \deq
\text{argmax}_{a'}A^\pi_\deltat(s, a')$
are the unique solution to the Bellman and consistency equations
\begin{align}
	V^{\pi}_\deltat(s) + \deltat\, A^{\pi}_\deltat(s, a) &=
	r\,\deltat + \gamma^{\deltat}  \,\E_{s'} V^{\pi}_\deltat(s')\label{eq:bellman_A}\\
	A^{\pi}_\deltat(s, \pi(s)) &= 0\label{eq:max_A}.
\end{align}
as seen in \ref{subsec:reparam}. Thus $V_{\theta}$
and $A_{\psi}$ are trained to approximately solve these equations.

Maximization over actions for $\pi$ is implemented exactly for discrete actions,
and, for continuous actions, approximated by a policy neural network $\pi_\phi(s)$ trained to maximize $A_\psi(s,
\pi_\phi(s))$, similarly to
\cite{ddpg}.

Eq.~\eqref{eq:max_A}
is directly verified by $A_{\psi}$, owing to the reparametrization
$A_\psi(s, a) = \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s))$, described
in~\ref{subsec:reparam}.  To approximately verify~\eqref{eq:bellman_A}, the
corresponding squared Bellman residual is minimized by an approximate gradient descent.
The update equations when learning from a transition $(s, a, r, s')$, either from
an exploratory trajectory or from a replay buffer~\cite{dqn}, are
\begin{align}
	\delta Q_\deltat &\leftarrow A_\psi(s, a)\, \deltat - \left(r
	\,\deltat + \gamma^{\deltat}\, V_\theta(s') - V_\theta(s)\right)
	\label{eq:approx_deltaQ}\\
	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat
	\,\partial_{\theta} V_\theta(s) \,\frac{\delta Q_\deltat}{\deltat}
	\label{eq:approx_bellman_V}\\
	\psi_\deltat &\leftarrow \psi_\deltat + \eta^A_\deltat
	\,\partial_{\psi} A_\psi(s, a) \,\frac{\delta Q_\deltat}{\deltat}.
	\label{eq:approx_bellman_A}
\end{align} 
where the $\eta$'s are learning rates.
Appropriate scalings for the learning rates $\eta^V_\deltat$ and
$\eta^A_\deltat$ in terms of $\deltat$ to obtain a well defined continuous
limit are derived next.
% As we will see in the following section,
% to obtain a well defined continuous time limit, the learning rates can be rewritten as
% $\eta^i_\deltat = \alpha^i \deltat$, where $\alpha^1$ and $\alpha^2$ are independant
% of $\deltat$.

@@

Informally, in the parameter updates
\eqref{eq:approx_deltaQ}--\eqref{eq:approx_bellman_A}, the quantity $\delta
Q_\deltat$ is of order $O(\deltat)$, because $s'=s+O(\deltat)$ in a
near-continuous system. Therefore, $\delta
Q_\deltat/\deltat$ is $O(1)$, so that the gradients used to
update $\theta$ and $\psi$ are $O(1)$. Therefore, if the
learning rates are of order $\deltat$, one would expect 
the parameters $\theta$ and $\psi$ to change by $O(\deltat)$ in
each time interval $\deltat$, thus hopefully converging to smooth
continuous-time trajectories. The next theorem formally confirms that
learning rates of order $\deltat$ are the only possibility.

### 2019-01-30 paper/icml_drau/discussions.tex

@@

\paragraph{Knowing $\deltat$.} Applying Alg.~\ref{alg:dau} requires
knowledge of the timestep $\deltat$. In most environments, this is
readily available, or even directly set by the practitioner: depending on
the environment it is given by the frame rate, the maximum frequency of
actuators or observation acquisition, etc.

>>>>

Note that Alg.~\ref{alg:dau} requires
knowledge of the timestep $\deltat$. In most environments, this is
readily available, or even directly set by the practitioner: depending on
the environment it is given by the frame rate, the maximum frequency of
actuators or observation acquisition, the timestep of a physics
simulator, etc.

@@

### 2019-01-30 paper/icml_drau/supplementary.tex
.

We now give proofs for all the results presented in the paper. Most
proofs follow standard patterns from calculus and numerical schemes for
differential equations, except for Theorem~\ref{thm:policyimprovement},
which uses an argument specific to reinforcement learning to prove that
the continuous-time advantage function contains all the necessary
information for policy values.

@@

In what follows, we assume that the continuous-time reward function $r\from {\cal S} \times {\cal A} \rightarrow \bb{R}$
is bounded, to ensure existence of $V^\pi$ and $V^\pi_\deltat$ for all $\deltat$
.\begin{theorem}
	Assume that $r\from {\cal S} \times {\cal A} \rightarrow \bb R$ is bounded, and
	that $s \rightarrow r(s, \pi(s))$ is $L_r$-Lipschitz continuous, then
	for all $s \in {\cal S}$, one has
	%\begin{equation}
	%\label{eq:conv-value}
	$V^\pi_\deltat(s) = V^\pi(s) + \smallo(\deltat)$
	%\end{equation}
	when $\deltat\to 0$.
	\label{th:conv-value}
\end{theorem}
\begin{proof}
	\input{proof_conv_value.tex}
\end{proof}

>>>>

We now give proofs for all the results presented in the paper. Most
proofs follow standard patterns from calculus and numerical schemes for
differential equations, except for Theorem~\ref{thm:policyimprovement},
which uses an argument specific to reinforcement learning to prove that
the continuous-time advantage function contains all the necessary
information for policy improvement.

@@

In what follows, we assume that the continuous-time reward function $r\from {\cal S} \times {\cal A} \rightarrow \bb{R}$
is bounded, to ensure existence of $V^\pi$ and $V^\pi_\deltat$ for all $\deltat$
.\begin{theorem}
	Assume that $r\from {\cal S} \times {\cal A} \rightarrow \bb R$ is bounded, and
	that $s \rightarrow r(s, \pi(s))$ is $L_r$-Lipschitz continuous, then
	for all $s \in {\cal S}$, one has
	%\begin{equation}
	%\label{eq:conv-value}
	$V^\pi_\deltat(s) = V^\pi(s) + \smallo(1)$
	%\end{equation}
	when $\deltat\to 0$.
	\label{th:conv-value}
\end{theorem}
\begin{proof}
	\input{proof_conv_value.tex}
\end{proof}
